A Heuristic Approach towards Drawings of Graphs with High Crossing Resolution

### 2019-07-25 paper/journal.tex
Almost done with reviewer 1. One more comment to be addressed.

\usepackage{amsmath,amssymb}
\usepackage{graphicx,color,pgfplots}
\usepackage[font=small]{subfig,caption}
\usepackage{todonotes,lineno,paralist,soul}
\usepackage{rotating,hyperref,multicol}
\usepackage{booktabs,tabularx,colortbl}
\graphicspath{{figures/}}

@@

In this paper, we propose a new heuristic algorithm for the crossing resolution maximization problem and we experimentally compare it against the known approaches from the literature. Our experimental evaluation indicates that the new heuristic produces drawings with better crossing resolution, but this comes at the cost of slightly higher aspect ratio, especially when the input graph is large. 
\end{abstract}

@@

\paragraph{Preliminaries:}
Unless otherwise specified, in this paper we consider simple undirected graphs. Let $G=(V,E)$ be such a graph. The degree of vertex $u\in V$ of $G$ is denoted by $d(u)$. The degree $d(G)$ of  graph $G$ is defined as the maximum degree of its vertices, i.e., $d(G)=\max_{u\in V}d(u)$. Given a drawing $\Gamma(G)$ of $G$, we denote by $p(u)=(x_u,y_u)$ the position of vertex $u \in V$ of $G$ in $\Gamma(G)$.

\begin{table*}
\caption{Summary of the results for the Graph Drawing Contest 2017.}
\centering
\begin{tabular}{c|c|c|c|c|c|c}
\toprule
\rowcolor{gray!30!}~Graph~ & ~no. of vertices~ & ~no. of edges~ & ~CoffeeVM~ & ~T\"ubingenMidnight~ & ~Time restricted~ & ~Our best~ \\
\midrule
\rowcolor{gray!30!} 1  &  9  & 18    & 90$^\circ$ & 77$^\circ$ & 90$^\circ$ & 90$^\circ$\\
\rowcolor{gray!00!} 2  & 12  & 36    & 88$^\circ$ & 42$^\circ$ & 88$^\circ$ & 89$^\circ$\\
\rowcolor{gray!30!} 3  & 16  & 42    & 90$^\circ$ & 89$^\circ$ & 88$^\circ$ & 90$^\circ$\\
\rowcolor{gray!00!} 4  & 17  & 54    & 89$^\circ$ & 89$^\circ$ & 77$^\circ$ & 89$^\circ$\\
\rowcolor{gray!30!} 5  & 30  & 95    & 80$^\circ$ & 30$^\circ$ & 79$^\circ$ & 87$^\circ$\\
\rowcolor{gray!00!} 6  & 40  & 75    & 90$^\circ$ & 78$^\circ$ & 90$^\circ$ & 90$^\circ$\\
\rowcolor{gray!30!} 7  & 80  & 254   & 57$^\circ$ & 34$^\circ$ & 56$^\circ$ & 64$^\circ$\\
\rowcolor{gray!00!} 8  & 85  & 136   & 85$^\circ$ & 61$^\circ$ & 81$^\circ$ & 89$^\circ$\\
\rowcolor{gray!30!} 9  & 149 & 258   & 60$^\circ$ & 9$^\circ$  & 55$^\circ$ & 88$^\circ$\\
\rowcolor{gray!00!} 10 & 174 & 344   & 21$^\circ$ & 4$^\circ$  & 24$^\circ$ & 24$^\circ$\\
\rowcolor{gray!30!} 11 & 250 & 400   & 47$^\circ$ & 6$^\circ$  & 57$^\circ$ & 72$^\circ$\\
\rowcolor{gray!00!} 12 & 300 & 987   & 36$^\circ$ & 5$^\circ$  & 26$^\circ$ & 36$^\circ$\\
\rowcolor{gray!30!} 13 & 392 & 562   & 25$^\circ$ & 4$^\circ$  & 22$^\circ$ & 34$^\circ$\\
\rowcolor{gray!00!} 14 & 510 & 820   & 34$^\circ$ & 5$^\circ$  & 30$^\circ$ & 43$^\circ$\\
\rowcolor{gray!30!} 15 & 650 & 2,173 & 21$^\circ$ & 4$^\circ$  & 13$^\circ$ & 29$^\circ$\\
\rowcolor{gray!00!} 16 & 998 & 1,596 & 20$^\circ$ & 3$^\circ$  & 12$^\circ$ & 26$^\circ$\\
\bottomrule
\end{tabular}
\label{tab:gdContest2017}
\end{table*}

@@

It is far from clear that the wining team was the team  ``\texttt{CoffeeVM}''; details for their approach can be found in~\cite{DBLP:journals/corr/abs-1807-09483}. Our team ``\texttt{TuebingenMidnight}''  miserably performed; we actually took the third position out of a total of four teams that participated. As already mentioned, each team had only one hour to compute a layout for each of the challenge graphs, which in year 2017 were in total $16$. In Table~\ref{tab:gdContest2017}, we additionally provide results that were achieved by the algorithm presented in this paper in two settings; within the time limit of the contest (see column ``Time restricted''), as well as our best results, which were achieved without any time limitations (see column ``Our best'').

@@

\begin{table*}
\caption{Summary of the results for the Graph Drawing Contest 2018.}
\centering
\begin{tabular}{c|c|c|c|c|c|c}
\toprule
\rowcolor{gray!30!} Graph & no. of vertices & no. of edges  & Grid size & T\"ubingenColdShower & CoffeeVM+ & Our best \\
\midrule
\rowcolor{gray!30!} 1  & 10    & 32    & 100 $\times$ 100             & 87$^\circ$ & 58$^\circ$ & 88$^\circ$\\
\rowcolor{gray!00!} 2  & 12    & 24    & 100 $\times$ 100             & 90$^\circ$ & 90$^\circ$ & 90$^\circ$\\
\rowcolor{gray!30!} 3  & 18    & 49    & 150 $\times$ 150             & 88$^\circ$ & 85$^\circ$ & 90$^\circ$\\
\rowcolor{gray!00!} 4  & 23    & 46    & 200 $\times$ 200             & 88$^\circ$ & 60$^\circ$ & 89$^\circ$\\
\rowcolor{gray!30!} 5  & 30    & 91    & 200 $\times$ 200             & 71$^\circ$ & 48$^\circ$ & 78$^\circ$\\
\rowcolor{gray!00!} 6  & 40    & 68    & 200 $\times$ 200             & 89$^\circ$ & 71$^\circ$ & 90$^\circ$\\
\rowcolor{gray!30!} 7  & 50    & 94    & 250 $\times$ 250             & 80$^\circ$ & 43$^\circ$ & 83$^\circ$\\
\rowcolor{gray!00!} 8  & 102   & 400   & 10,000 $\times$ 10,000       & 34$^\circ$ & 27$^\circ$ & 43$^\circ$\\
\rowcolor{gray!30!} 9  & 128   & 256   & 10,000 $\times$ 10,000       & 77$^\circ$ & 60$^\circ$ & 83$^\circ$\\
\rowcolor{gray!00!} 10 & 500   & 1,534 & 30,000 $\times$ 30,000       & 13$^\circ$ & 19$^\circ$ & 24$^\circ$\\
\rowcolor{gray!30!} 11 & 709   & 1,602 & 50,000 $\times$ 50,000       & 11$^\circ$ & 22$^\circ$ & 18$^\circ$\\
\rowcolor{gray!00!} 12 & 1,800 & 6,961 & 100,000 $\times$ 100,000     & 74$^\circ$ &  9$^\circ$ & 80$^\circ$\\
\rowcolor{gray!30!} 13 & 1,500 & 4,467 & 1,000,000 $\times$ 1,000,000 &  2$^\circ$ &  2$^\circ$ & 4$^\circ$\\
\rowcolor{gray!00!} 14 & 3,000 & 5,899 & 1,000,000 $\times$ 1,000,000 &  3$^\circ$ & 11$^\circ$ & 11$^\circ$\\
\bottomrule
\end{tabular}
\label{tab:gdContest2018}
\end{table*}

@@

\paragraph{Structure of the paper:}
The remainder of this paper is structured as follows. 
%Section~\ref{sec:relatedwork} overviews related works. 
Our algorithm is presented in detail in Section~\ref{sec:algorithm} and is experimentally evaluated against the ones by Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13} and by Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13} in Section~\ref{sec:experiments}, where we also discuss our insights from this project. We conclude in Section~\ref{sec:conclusions} with a discussion of useful observations and insights that we obtained while working on this project.

@@

In this section, we describe our heuristic for obtaining drawings with high crossing resolution. The input of our heuristic consists of a graph $G$ and an initial drawing $\Gamma_0$ of $G$ with crossing resolution $c(\Gamma_0)$. We assume that no two edges of graph $G$ overlap in drawing $\Gamma_0$, i.e., $c(\Gamma_0)>0$. A circular drawing or a drawing obtained by applying a force-directed algorithm on graph $G$ clearly meets this precondition.

Our algorithm is iterative and at each iteration performs some operations that are mainly based on randomization. At the $i$-th iteration, we assume that we have computed a drawing $\Gamma_{i-1}$ of graph $G$ of crossing resolution $c(\Gamma_{i-1}) \geq c(\Gamma_0)$. In other words, we assume, as an invariant for our algorithm, that the crossing resolution cannot be decreased at some iteration of our algorithm (so, in a sense, our algorithm resembles probabilistic hill climbing approaches). We proceed by choosing a vertex of $\Gamma_{i-1}$ arbitrarily at random based on the content of the so-called \emph{vertex-pool}, which may contain:

@@

Intuitively, the critical vertices are the endpoints of the edges that define the crossing resolution of drawing $\Gamma_{i-1}$. To formally define them, we first need to introduce the notion of critical edge-pairs. A pair of edges $e$ and $e'$ is called \emph{critical} in drawing $\Gamma_{i-1}$ if and only if edges $e$ and $e'$ cross in $\Gamma_{i-1}$ and the minimum angle that is formed at their crossing point is equal to the crossing resolution of drawing $\Gamma_{i-1}$ (that is, to $c(\Gamma_{i-1})$). The set of critical vertices of drawing $\Gamma_{i-1}$ is then defined by the four endvertices of each critical edge-pair.

The role of critical vertices is central in our algorithm\footnote{If the focus is not on the critical vertices for a large graph, then our algorithm will need a large number of iterations to converge to a solution with good crossing resolution, because it is simply very unlikely to select to move one of the vertices that define the crossing resolution.}: By appropriately changing the location of a critical vertex or of a vertex that is in the neighbourhood of the critical vertices, we naturally expect to improve the crossing resolution of the current drawing. We turned this observation into an algorithmic implementation through a probabilistic random selection procedure, so that the vertices at graph-distance~$i$ from the ones of the vertex-pool have higher probability for selection than the corresponding ones at graph-distance~$j$  in $G$, when $0 \leq i<j$. In other words, the closer a vertex is to the critical vertices, the more likely it is to be chosen. On the other hand, if the vertex-pool contains all vertices of the graph, then each vertex can be chosen with the same probability.

@@

So far, we have described the main idea of our algorithm, which at each iteration chooses uniformly at random a vertex of the current drawing to move (based on the content of the vertex-pool), so to improve the crossing resolution. Next, we describe how to compute the new position of the chosen vertex in the next drawing $\Gamma_i$. Recall that the crossing resolution $c(\Gamma_{i})$ of drawing $\Gamma_{i}$ that is obtained after the $i$-th iteration of the algorithm must be at least as large as the crossing resolution $c(\Gamma_{i-1})$ of drawing $\Gamma_{i-1}$ (by the invariant of our algorithm), that is, $c(\Gamma_i) \ge c(\Gamma_{i-1})$.

@@

In general, aesthetically pleasant drawings of graphs are usually the result of compromising between different aesthetic criteria. Towards this direction, we discuss in this section interesting variants of our algorithm, which are motivated by the following observation that we made while working on this project (see Section~\ref{sec:conclusions}): Drawings that are optimised only in terms of the crossing resolution tend to have bad aspect ratio (that is, the ratio of the longest to the shortest edge is large) and poor angular resolution (that is, the angles formed by adjacent edges are small). The former seems to be a consequence of the fact that drawings with good crossing resolution tend to be quite demanding in area. For the latter, observe that if in a drawing all edges are either almost horizontally or almost vertically drawn, such that only ``horizontal'' and ``vertical'' edges cross, then the crossing resolution of this drawing is arbitrarily close to $90^\circ$, while its angular resolution is arbitrarily close to~$0^\circ$.

% ==================================================================
\myparagraph{Aspect ratio}
% ==================================================================
%
Formally, the aspect ratio of a drawing is the ratio of the length of its longest edge to the length of its shortest edge. Sometimes it is also used as a measure of the area of non-grid drawings. It was easy to instruct our algorithm to prevent producing drawings with aspect ratio either higher than the one of the starting layout or higher than a given input value. What we simply had to do was to reject candidate positions, which violate this precondition.

@@

The experiment was performed on a Linux laptop with four cores at 2.4 GHz and 8 GB RAM. The experiment was performed as follows. Initially, each graph of our expertiment was laid out using the SmartOrganic layouter of yFiles~\cite{DBLP:books/sp/04/WieseE004}. Starting from this layout, every graph was drawn with:
%
\begin{itemize}
\item[--] our algorithm with $\delta_{max}=\frac{1}{2}\max\{w,h\}$, where $w$ and $h$ are the width and the height of the initial drawing, respectively, $\delta_{min}=\frac{1}{100}\delta_{max}$ and $\rho=10$,
\item[--] our algorithm restricted not to violate the aspect ratio of the initial layout, 
\item[--] the force-directed algorithm by Argyriou et al., and
\item[--] the force-directed algorithm by Huang et al.
\end{itemize}

@@

\myparagraph{Crossing resolution} Our results for the crossing resolution are summarized in Fig.~\ref{fig:cr-res}. Here, each algorithm was adjusted to maximize exclusively the crossing resolution (i.e., by ignoring the drawing's angular resolution). It is immediate to see that our algorithm outperforms all other ones in terms of the crossing resolution of the produced drawings, when we do not impose any restriction on the aspect ratio of the computed drawings; refer to the solid-black curve, denoted as \emph{Unrestricted}, in Fig.~\ref{fig:cr-res-1}. The variant of our algorithm, which does not violate the aspect ratio of the initial layout, leads to drawings with slightly smaller crossing resolution; refer to the solid-gray curve, denoted as \emph{AR-restricted}, in Fig.~\ref{fig:cr-res-1}. Finally, the two force-directed algorithms seem to produce drawings with worse crossing resolution; refer to the dotted-gray and dotted-black curves of Fig.~\ref{fig:cr-res-1} (by Argyriou et al.\ and by Huang et al., respectively).

% ==================================================================
%\input{cr-res-rome.tex}
% ==================================================================
\begin{figure*}[t]
	\centering
	\subfloat[\label{fig:cr-res-1}{Crossing resolution vs no.~of vertices}]{
	\includegraphics[scale=0.99,page=1]{figures/rome}}
	\hfil
	\subfloat[\label{fig:cr-res-2}{Aspect ratio vs no.~of vertices}]{
	\includegraphics[scale=0.99,page=2]{figures/rome}}

@@

% ==================================================================
%\input{to-res-rome.tex}
% ==================================================================
\begin{figure*}[t]
	\centering
	\subfloat[\label{fig:to-res-1}{Crossing resolution vs no.~of vertices}]{
	\includegraphics[scale=0.99,page=5]{figures/rome}}
	\hfil
	\subfloat[\label{fig:to-res-2}{Aspect ratio vs no.~of vertices}]{
	\includegraphics[scale=0.99,page=6]{figures/rome}}

@@

% ==================================================================
%\input{an-res-rome.tex}
% ==================================================================
\begin{figure*}[t]
	\centering
	\subfloat[\label{fig:an-res-1}{Crossing resolution vs no.~of vertices}]{
	\includegraphics[scale=0.99,page=9]{figures/rome}}
	\hfil
	\subfloat[\label{fig:an-res-2}{Aspect ratio vs no.~of vertices}]{
	\includegraphics[scale=0.99,page=10]{figures/rome}}

@@

% ==================================================================
%\input{cr-res-north.tex}
% ==================================================================
\begin{figure*}[t!]
	\centering
	\subfloat[\label{fig:northCrossing-1}{Crossing resolution vs no.~of vertices}]{
	\includegraphics[scale=0.99,page=1]{figures/north}}
	\hfil
	\subfloat[\label{fig:northCrossing-2}{Aspect ratio vs no.~of vertices}]{
	\includegraphics[scale=0.99,page=2]{figures/north}}

@@

% ==================================================================
%\input{to-res-north.tex}
% ==================================================================
\begin{figure*}[t!]
	\centering
	\subfloat[\label{fig:northTotal-1}{Crossing resolution vs no.~of vertices}]{
	\includegraphics[scale=0.99,page=5]{figures/north}}
	\hfil
	\subfloat[\label{fig:northTotal-2}{Aspect ratio vs no.~of vertices}]{
	\includegraphics[scale=0.99,page=6]{figures/north}}

@@

% ==================================================================
%\input{an-res-north.tex}
% ==================================================================
\begin{figure*}[t!]
	\centering
	\subfloat[\label{fig:northAngular-1}{Crossing resolution vs no.~of vertices}]{
	\includegraphics[scale=0.99,page=9]{figures/north}}
	\hfil
	\subfloat[\label{fig:northAngular-2}{Aspect ratio vs no.~of vertices}]{
	\includegraphics[scale=0.99,page=10]{figures/north}}

@@

\section{Sample Drawings}
\label{app:samples}

>>>>

\usepackage{amsmath,amssymb,adjustbox}
\usepackage{graphicx,color,pgfplots}
\usepackage[font=small]{subfig,caption}
\usepackage{todonotes,lineno,paralist,soul}
\usepackage{rotating,hyperref,multicol}
\usepackage{booktabs,tabularx,colortbl}
\graphicspath{{figures/}}

@@

In this paper, we propose a new heuristic algorithm for the crossing resolution maximization problem and we experimentally compare it against the known approaches from the literature. Our experimental evaluation indicates that the new heuristic produces drawings with better crossing resolution, but this comes at the cost of slightly higher edge-length ratio, especially when the input graph is large. 
\end{abstract}

@@

\paragraph{Preliminaries:}
In this paper we consider simple undirected graphs. Let $G=(V,E)$ be such a graph. The degree of vertex $u\in V$ of $G$ is denoted by $d(u)$. The degree $d(G)$ of  graph $G$ is defined as the maximum degree of its vertices, i.e., $d(G)=\max_{u\in V}d(u)$. Given a drawing $\Gamma(G)$ of $G$, we denote by $p(u)=(x_u,y_u)$ the position of vertex $u \in V$ of $G$ in $\Gamma(G)$.

\begin{table*}
\caption{Summary of the results for the Graph Drawing Contest 2017; the last two columns correspond to results presented in this paper (with and without time limitations, respectively).}
\centering
\begin{adjustbox}{width=1\textwidth}
\begin{tabular}{c|c|c|c|c|c|c}
\toprule
\rowcolor{gray!30!}~Graph~ & ~no.~of vertices~ & ~no.~of edges~ & CoffeeVM & T\"ubingenMidnight & Time restricted & No-time restriction \\
\midrule
\rowcolor{gray!30!} 1  &  9  & 18    & 90$^\circ$ & 77$^\circ$ & 90$^\circ$ & 90$^\circ$\\
\rowcolor{gray!00!} 2  & 12  & 36    & 88$^\circ$ & 42$^\circ$ & 88$^\circ$ & 89$^\circ$\\
\rowcolor{gray!30!} 3  & 16  & 42    & 90$^\circ$ & 89$^\circ$ & 88$^\circ$ & 90$^\circ$\\
\rowcolor{gray!00!} 4  & 17  & 54    & 89$^\circ$ & 89$^\circ$ & 77$^\circ$ & 89$^\circ$\\
\rowcolor{gray!30!} 5  & 30  & 95    & 80$^\circ$ & 30$^\circ$ & 79$^\circ$ & 87$^\circ$\\
\rowcolor{gray!00!} 6  & 40  & 75    & 90$^\circ$ & 78$^\circ$ & 90$^\circ$ & 90$^\circ$\\
\rowcolor{gray!30!} 7  & 80  & 254   & 57$^\circ$ & 34$^\circ$ & 56$^\circ$ & 64$^\circ$\\
\rowcolor{gray!00!} 8  & 85  & 136   & 85$^\circ$ & 61$^\circ$ & 81$^\circ$ & 89$^\circ$\\
\rowcolor{gray!30!} 9  & 149 & 258   & 60$^\circ$ & 9$^\circ$  & 55$^\circ$ & 88$^\circ$\\
\rowcolor{gray!00!} 10 & 174 & 344   & 21$^\circ$ & 4$^\circ$  & 24$^\circ$ & 24$^\circ$\\
\rowcolor{gray!30!} 11 & 250 & 400   & 47$^\circ$ & 6$^\circ$  & 57$^\circ$ & 72$^\circ$\\
\rowcolor{gray!00!} 12 & 300 & 987   & 36$^\circ$ & 5$^\circ$  & 26$^\circ$ & 36$^\circ$\\
\rowcolor{gray!30!} 13 & 392 & 562   & 25$^\circ$ & 4$^\circ$  & 22$^\circ$ & 34$^\circ$\\
\rowcolor{gray!00!} 14 & 510 & 820   & 34$^\circ$ & 5$^\circ$  & 30$^\circ$ & 43$^\circ$\\
\rowcolor{gray!30!} 15 & 650 & 2,173 & 21$^\circ$ & 4$^\circ$  & 13$^\circ$ & 29$^\circ$\\
\rowcolor{gray!00!} 16 & 998 & 1,596 & 20$^\circ$ & 3$^\circ$  & 12$^\circ$ & 26$^\circ$\\
\bottomrule
\end{tabular}
\end{adjustbox}
\label{tab:gdContest2017}
\end{table*}

@@

It is far from clear that the wining team was the team  ``\texttt{CoffeeVM}''; details for their approach can be found in~\cite{DBLP:journals/corr/abs-1807-09483}. Our team ``\texttt{TuebingenMidnight}''  performed miserably; we actually took the third position out of a total of four teams that participated. As already mentioned, each team had only one hour to compute a layout for each of the challenge graphs, which in year 2017 were in total $16$. In Table~\ref{tab:gdContest2017}, we additionally provide results that were achieved by the algorithm presented in this paper in two settings; within the time limit of the contest (see column ``Time restricted''), as well as s results, which were achieved without any time limitations (see column ``No-time restriction'').

@@

\begin{table*}
\caption{Summary of the results for the Graph Drawing Contest 2018.}
\centering
\begin{adjustbox}{width=1\textwidth}
\begin{tabular}{c|c|c|c|c|c|c}
\toprule
\rowcolor{gray!30!} Graph & no.~of vertices & no.~of edges  & Grid size & T\"ubingenColdShower & CoffeeVM+ & No-time restriction \\
\midrule
\rowcolor{gray!30!} 1  & 10    & 32    & 100 $\times$ 100             & 87$^\circ$ & 58$^\circ$ & 88$^\circ$\\
\rowcolor{gray!00!} 2  & 12    & 24    & 100 $\times$ 100             & 90$^\circ$ & 90$^\circ$ & 90$^\circ$\\
\rowcolor{gray!30!} 3  & 18    & 49    & 150 $\times$ 150             & 88$^\circ$ & 85$^\circ$ & 90$^\circ$\\
\rowcolor{gray!00!} 4  & 23    & 46    & 200 $\times$ 200             & 88$^\circ$ & 60$^\circ$ & 89$^\circ$\\
\rowcolor{gray!30!} 5  & 30    & 91    & 200 $\times$ 200             & 71$^\circ$ & 48$^\circ$ & 78$^\circ$\\
\rowcolor{gray!00!} 6  & 40    & 68    & 200 $\times$ 200             & 89$^\circ$ & 71$^\circ$ & 90$^\circ$\\
\rowcolor{gray!30!} 7  & 50    & 94    & 250 $\times$ 250             & 80$^\circ$ & 43$^\circ$ & 83$^\circ$\\
\rowcolor{gray!00!} 8  & 102   & 400   & 10,000 $\times$ 10,000       & 34$^\circ$ & 27$^\circ$ & 43$^\circ$\\
\rowcolor{gray!30!} 9  & 128   & 256   & 10,000 $\times$ 10,000       & 77$^\circ$ & 60$^\circ$ & 83$^\circ$\\
\rowcolor{gray!00!} 10 & 500   & 1,534 & 30,000 $\times$ 30,000       & 13$^\circ$ & 19$^\circ$ & 24$^\circ$\\
\rowcolor{gray!30!} 11 & 709   & 1,602 & 50,000 $\times$ 50,000       & 11$^\circ$ & 22$^\circ$ & 18$^\circ$\\
\rowcolor{gray!00!} 12 & 1,800 & 6,961 & 100,000 $\times$ 100,000     & 74$^\circ$ &  9$^\circ$ & 80$^\circ$\\
\rowcolor{gray!30!} 13 & 1,500 & 4,467 & 1,000,000 $\times$ 1,000,000 &  2$^\circ$ &  2$^\circ$ & 4$^\circ$\\
\rowcolor{gray!00!} 14 & 3,000 & 5,899 & 1,000,000 $\times$ 1,000,000 &  3$^\circ$ & 11$^\circ$ & 11$^\circ$\\
\bottomrule
\end{tabular}
\end{adjustbox}
\label{tab:gdContest2018}
\end{table*}

@@

\paragraph{Structure of the paper:}
The remainder of this paper is structured as follows. 
%Section~\ref{sec:relatedwork} overviews related works. 
Our algorithm is presented in detail in Section~\ref{sec:algorithm} and is experimentally evaluated against the ones by Argyriou et al.~\cite{DBLP:journals/cj/ArgyriouBS13} and by Huang et al.~\cite{DBLP:journals/vlc/HuangEHL13} in Section~\ref{sec:experiments}, where we also discuss our insights from this project. In Section~\ref{sec:samples} we present sample drawings produced by our algorithm. We conclude in Section~\ref{sec:conclusions} with a discussion of useful observations and insights that we obtained while working on this project.

@@

In this section, we describe our heuristic for obtaining drawings with high crossing resolution. The input of our heuristic consists of a graph $G$ and an initial drawing $\Gamma_0$ of $G$ with crossing resolution $c(\Gamma_0)$. We assume that no two edges of graph $G$ overlap in drawing $\Gamma_0$, i.e., $c(\Gamma_0)>0$. A circular drawing or a drawing obtained by applying a force-directed algorithm on graph $G$ clearly meets this precondition. The output of our heuristic will be a (not necessarily grid) drawing $\Gamma$ with crossing resolution $c(\Gamma)$, such that $c(\Gamma) \geq c(\Gamma_0)$. In Section~\ref{ssec:variants}, we discuss how one can easily modify our heuristic so that $\Gamma$ is necessarily a grid drawing, if  the input drawing $\Gamma_0$ is a grid drawing, as well.

Our algorithm is iterative and at each iteration performs some operations that are mainly based on randomization. At the $i$-th iteration, we assume that we have computed a drawing $\Gamma_{i-1}$ of graph $G$ with crossing resolution $c(\Gamma_{i-1})$. As an invariant for our algorithm, we further assume that for all values of $k$ with $0 \leq k \leq i-1$ the following condition holds %
%
$$c(\Gamma_{k-1}) \leq c(\Gamma_k).$$ 
%
In other words, we assume that the crossing resolution cannot be decreased at some iteration of our algorithm (so, in a sense, our algorithm resembles probabilistic hill climbing approaches), which also implies that $c(\Gamma_{i-1}) \geq c(\Gamma_{0})$. We proceed by choosing a vertex of $\Gamma_{i-1}$ arbitrarily at random based on the content of the so-called \emph{vertex-pool}, which may contain:

@@

Intuitively, the critical vertices are the endpoints of the edges that define the crossing resolution of drawing $\Gamma_{i-1}$. To formally define them, we first need to introduce the notion of critical edge-pairs. A pair of edges $e$ and $e'$ is called \emph{critical} in drawing $\Gamma_{i-1}$ if and only if edges $e$ and $e'$ cross in $\Gamma_{i-1}$ and the minimum angle that is formed at their crossing point is equal to the crossing resolution of drawing $\Gamma_{i-1}$ (that is, to $c(\Gamma_{i-1})$). The set of critical vertices of drawing $\Gamma_{i-1}$ is then defined by the four endvertices of each critical edge-pair. It is worth noting at this point that several well-known problems related to precision in real-number computations may occur, while computing the set of critical vertices of drawing $\Gamma_{i-1}$. Here, we assume that this set has been computed correctly up to the reliability of computations in the used machine.

The role of critical vertices is central in our algorithm\footnote{If the focus is not on the critical vertices for a large graph, then our algorithm will need a large number of iterations to converge to a solution with good crossing resolution, because it is simply very unlikely to select to move one of the vertices that define the crossing resolution.}: By appropriately changing the location of a critical vertex or of a vertex that is in the neighbourhood of the critical vertices, we naturally expect to improve the crossing resolution of the current drawing. We turned this observation into an algorithmic implementation through a probabilistic random selection procedure, so that the vertices at graph-distance~$i$ from the ones of the vertex-pool have higher probability for selection than the corresponding ones at graph-distance~$j$  in $G$, when $0 \leq i<j$. In other words, the closer a vertex is to the critical vertices, the more likely it is to be chosen\footnote{Note that in approach we have chosen the graph-distance as the criterion for choosing the vertices to be moved, mainly for reasons of efficiency. Alternatively, one could use the Euclidean distance for this purpose.}. On the other hand, if the vertex-pool contains all vertices of the graph, then each vertex can be chosen with the same probability.

@@

So far, we have described the main idea of our algorithm, which at each iteration chooses  a vertex of the current drawing to move (based on the content of the vertex-pool), so to improve the crossing resolution. Next, we describe how to compute the new position of the chosen vertex in the next drawing $\Gamma_i$. Recall that the crossing resolution $c(\Gamma_{i})$ of drawing $\Gamma_{i}$ that is obtained after the $i$-th iteration of the algorithm must be at least as large as the crossing resolution $c(\Gamma_{i-1})$ of drawing $\Gamma_{i-1}$ (by the invariant of our algorithm), that is, $c(\Gamma_i) \ge c(\Gamma_{i-1})$.

@@

In general, aesthetically pleasant drawings of graphs are usually the result of compromising between different aesthetic criteria. Towards this direction, we discuss in this section interesting variants of our algorithm, which are motivated by the following observation that we made while working on this project (see Section~\ref{sec:conclusions}): Drawings that are optimised only in terms of the crossing resolution tend to have bad edge-length ratio (that is, the ratio of the longest to the shortest edge is large) and poor angular resolution (that is, the angles formed by adjacent edges are small). The former seems to be a consequence of the fact that drawings with good crossing resolution tend to be quite demanding in area. For the latter, observe that if in a drawing all edges are either almost horizontally or almost vertically drawn, such that only ``horizontal'' and ``vertical'' edges cross, then the crossing resolution of this drawing is arbitrarily close to $90^\circ$, while its angular resolution is arbitrarily close to~$0^\circ$.

% ==================================================================
\myparagraph{Edge-length ratio}
% ==================================================================
%
Formally, the edge-length ratio of a drawing is the ratio of the length of its longest edge to the length of its shortest edge; see, e.g.,~\cite{DBLP:journals/tcs/LazardLL19}. Sometimes it is also used as a measure of the area of non-grid drawings. It was easy to instruct our algorithm to prevent producing drawings with edge-length ratio either higher than the one of the starting layout or higher than a given input value. What we simply had to do was to reject candidate positions, which violate this precondition.

@@

The experiment was performed on a Linux laptop with four cores at 2.4 GHz and 8 GB RAM. The experiment was performed as follows. Initially, each graph of our expertiment was laid out using the SmartOrganic layouter of yFiles~\cite{DBLP:books/sp/04/WieseE004}. Starting from this layout, every graph was drawn with:
%
\begin{itemize}
\item[--] our algorithm with $\delta_{max}=\frac{1}{2}\max\{w,h\}$, where $w$ and $h$ are the width and the height of the initial drawing, respectively, $\delta_{min}=\frac{1}{100}\delta_{max}$ and $\rho=10$,
\item[--] our algorithm restricted not to violate the edge-length ratio of the initial layout, 
\item[--] the force-directed algorithm by Argyriou et al., and
\item[--] the force-directed algorithm by Huang et al.
\end{itemize}

@@

\myparagraph{Crossing resolution} Our results for the crossing resolution are summarized in Fig.~\ref{fig:cr-res}. Here, each algorithm was adjusted to maximize exclusively the crossing resolution (i.e., by ignoring the drawing's angular resolution). It is immediate to see that our algorithm outperforms all other ones in terms of the crossing resolution of the produced drawings, when we do not impose any restriction on the edge-length ratio of the computed drawings; refer to the solid-black curve, denoted as \emph{Unrestricted}, in Fig.~\ref{fig:cr-res-1}. The variant of our algorithm, which does not violate the edge-length ratio of the initial layout, leads to drawings with slightly smaller crossing resolution; refer to the solid-gray curve, denoted as \emph{ELR-restricted}, in Fig.~\ref{fig:cr-res-1}. Finally, the two force-directed algorithms seem to produce drawings with worse crossing resolution; refer to the dotted-gray and dotted-black curves of Fig.~\ref{fig:cr-res-1} (by Argyriou et al.\ and by Huang et al., respectively).

% ==================================================================
%\input{cr-res-rome.tex}
% ==================================================================
\begin{figure*}[t]
	\centering
	\subfloat[\label{fig:cr-res-1}{Crossing resolution vs no.~of vertices}]{
	\includegraphics[scale=0.99,page=1]{figures/rome}}
	\hfil
	\subfloat[\label{fig:cr-res-2}{Edge-length ratio vs no.~of vertices}]{
	\includegraphics[scale=0.99,page=2]{figures/rome}}

@@

% ==================================================================
%\input{to-res-rome.tex}
% ==================================================================
\begin{figure*}[t]
	\centering
	\subfloat[\label{fig:to-res-1}{Total resolution vs no.~of vertices}]{
	\includegraphics[scale=0.99,page=5]{figures/rome}}
	\hfil
	\subfloat[\label{fig:to-res-2}{Edge-length ratio vs no.~of vertices}]{
	\includegraphics[scale=0.99,page=6]{figures/rome}}

@@

% ==================================================================
%\input{an-res-rome.tex}
% ==================================================================
\begin{figure*}[t]
	\centering
	\subfloat[\label{fig:an-res-1}{Angular resolution vs no.~of vertices}]{
	\includegraphics[scale=0.99,page=9]{figures/rome}}
	\hfil
	\subfloat[\label{fig:an-res-2}{Edge-length ratio vs no.~of vertices}]{
	\includegraphics[scale=0.99,page=10]{figures/rome}}

@@

% ==================================================================
%\input{cr-res-north.tex}
% ==================================================================
\begin{figure*}[t!]
	\centering
	\subfloat[\label{fig:northCrossing-1}{Crossing resolution vs no.~of vertices}]{
	\includegraphics[scale=0.99,page=1]{figures/north}}
	\hfil
	\subfloat[\label{fig:northCrossing-2}{Edge-length ratio vs no.~of vertices}]{
	\includegraphics[scale=0.99,page=2]{figures/north}}

@@

% ==================================================================
%\input{to-res-north.tex}
% ==================================================================
\begin{figure*}[t!]
	\centering
	\subfloat[\label{fig:northTotal-1}{Total resolution vs no.~of vertices}]{
	\includegraphics[scale=0.99,page=5]{figures/north}}
	\hfil
	\subfloat[\label{fig:northTotal-2}{Edge-length ratio vs no.~of vertices}]{
	\includegraphics[scale=0.99,page=6]{figures/north}}

@@

% ==================================================================
%\input{an-res-north.tex}
% ==================================================================
\begin{figure*}[t!]
	\centering
	\subfloat[\label{fig:northAngular-1}{Angular resolution vs no.~of vertices}]{
	\includegraphics[scale=0.99,page=9]{figures/north}}
	\hfil
	\subfloat[\label{fig:northAngular-2}{Edge-length ratio vs no.~of vertices}]{
	\includegraphics[scale=0.99,page=10]{figures/north}}

@@

\section{Sample Drawings}
\label{sec:samples}
